---
title: "EXPLORATORY DATA ANALYSIS CSC 3220 - Real Estate Data"
author: |
  | Robert Bingham
  | Phonethep Nakhonekhong
  | Eli Parker
  | John Taylor
  | Johnathan Rich
date: "2022-11-08"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes: \usepackage{fvextra} \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

## Introduction

Our group, 5 little minds, have decided to look at real estate market data from 2012 to 2021 in order to predict future trends in the next 5 years. This data includes median list price for a given neighborhood in each state; median prices based upon housing unit categories, such as apartments, single-family housing, and condos; and year-to-year increases in sale prices of each unit.

### Format the data

For the purposes of this assignment, were are turning off all warnings and centering each graph.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
```

### Import the necessary libraries

```{r}
library("ggplot2")
library("DT")
library("pander")
library("corrplot")
library("zoo")
library("reshape")
library("scales")
library("tidyverse")
```

\newpage

### Import the Data

We decided to use the data from this url from Kaggle for our dataset: [\textcolor{blue}{Link}](https://www.kaggle.com/datasets/thuynyle/redfin-housing-market-data?select=state_market_tracker.tsv000)

```{r}
state_market.df <- read.table("../data/state_market_tracker.tsv000", sep = '\t', header = TRUE)
```

### Data Manipulation

Here, we have made R recognize  the variables in the dataset that pertain to specific days, (i.e, 9/21/2022) as actual dates using the built-in as.Date function. We have also divided the median sale price and list price of homes in each neighborhood by 1000 in order to  make the data more readable in subsequent graphs. There is no missing values in the dataset, so we do not have to do anything with that.

```{r}
state_market.df$period_begin <- as.Date(state_market.df$period_begin)
state_market.df$period_end <- as.Date(state_market.df$period_end)
state_market.df$median_sale_price <- state_market.df$median_sale_price / 1000
state_market.df$median_list_price <- state_market.df$median_list_price / 1000
state_market.df$property_type[state_market.df$property_type == "Multi-Family (2-4 Unit)"] <- "Multi-Family"
```

Out of the 48 variables in the dataset, these are the ones we will consider (the important variables) in our subsequent exploratory analysis:

* median sale price - **Median sale price the housing units of each of the 27,054 neighborhoods**
* median list price - **Median list price the housing units of each of the 27,054 neighborhoods**
* median sale price year-on-year - **Year-to-year percentage increase in median sale price of each neighborhood**
* median list price year-on-year - **Year-to-year percentage increase in median list price of each neighborbood**
* median sale ppsf - **Median sale price per-square-foot of each neighborhood**
* median list ppsf - **Median sale price per-square-foot of each neighborhood**
* homes sold year-old-year - **Increase in number of homes sold in each neighborhood year-on-year**
* pending sales year-on-year - **Increase in number of pending sales in each neighborhood year-on-year**
* new listings year-on-year - **Increase in number of new listings in each neighborhood year-on-year**
* inventory year-on-year - **Increase in number of all listings in each neighborhood year-on-year**
* sold_above_list_yoy – **Increase in number of homes sold above list price in each neighborhood**
* avg_sale_to_list_yoy – **Change in the ratio of homes sold to homes listed in each neighborhood**
* region – **Categorical variable  that classifies which region the neighborhood is in**
* state_code – **Categorical variable that is abbreviated state name**
* property_type – **Type of home**

\newpage

## Interpreting the Data

Here, we created a boxplot graph for the median list price of the homes in each neighborhood in the dataset in the year 2021, with each boxplot representing each state. Hawaii and and Connecticut were had the highest number of outliers. Since there were more outliers above the boxplot than below it in both cases, we attributed these anomalies to the high cost of living in both states, Since Hawaii is a vacation destination and Connecticut is a New England state with a close proximity to New York City. We can see that states with much lower overall median sale price, such as Oklahoma and Ohio, also have much lower variance. There is sarious skewness and interquartile ranges between each state. Some are skewed left, and some are skewed right. Majority have small interquartile ranges. A few have large interquartile ranges. Initially, median_sale_price is our target variable, since we want to learn how much money we can make from selling houses!

```{r, fig.align = 'center', fig.width = 12, fig.height = 10}
split_by_year <- split(state_market.df, format(state_market.df$period_begin, "%Y"))
ggplot(split_by_year[[length(split_by_year)]], aes(x = state_code, y = median_sale_price, color=median_sale_price)) +
  geom_boxplot(colour="#003366", fill="#66FFFF", alpha=5) +
  geom_point() +
  xlab("2021") +
  ylab("Median Sales Price (in thousands)") +
  ggtitle("Distribution of Median Sale Price per State") +
  theme(plot.title = element_text(hjust = 0.5))
```

\newpage

We created 2 histograms. One is for the median list price of housing units in each of the 27,054 neighborhoods from 2012 to 2021. The other is the median sale price of over the same length of time. We noticed that both graphs are skewed to the right, so we can assume that there are more homes sold and listed above the overall median home price than below it in the U.S. We also noticed that there is much less variance in the sale price than in the list price, with the right skew in the list price being much less pronounced in the list price than the sale price. Therefore, we can assume that if the asking/list price is higher than the median, the seller receives less in the final sale price in most cases.

```{r}
mean_years <- data.frame(state_market.df$period_begin, state_market.df$median_sale_price_yoy, state_market.df$state)

new_df <- subset.data.frame(state_market.df, select = c(state_code, median_list_price, median_sale_price),  drop = FALSE)

print(
    ggplot(new_df, aes_string(x=new_df$median_list_price))
    + geom_histogram(
      colour="darkorchid4", fill="darkorchid1", position="identity", bins=30, alpha=0.2
    )
    + ggtitle(paste("Frequency Distribution of median list price", sep=""))
    + theme(plot.title=element_text(hjust = 0.5))
    + xlab("Median List Price of Each Neighboorhood (in thousands)")
    + ylab("Frequency of Neighborhoods")
    + ylim(0, 12500))
```

```{r}
print(
    ggplot(new_df, aes_string(x=new_df$median_sale_price))
    + geom_histogram(
      colour="red", fill="firebrick1", position="identity", bins=30, alpha=0.2
    )
    + ggtitle(paste("Frequency Distribution of median sale price", sep=""))
    + theme(plot.title=element_text(hjust = 0.5))
    + xlab("Median Sale Price of Each Neighboorhood (in thousands)")
    + ylab("Frequency of Neighborhoods")
    + ylim(0, 12500))
```

\newpage

We can create a bin plot to demonstrate that home sale prices tend to aggregate below a million, between 100 thousand to 300 thousand. We have fitted a best fit line to show that there is a positive increase in median sales price over time, however, with a r-squared value of 0.125, only 12.5% of this increase can be explained by time. This is understandable, since the value of a home includes many factors such as: location, land, size, time built, etc.

```{r, fig.width = 12, fig.height = 10}
lm_eqn <- function(df){
    m <- lm(median_sale_price ~ as.yearmon(period_begin), df);
    eq <- substitute(italic(median_sale_price) == b %.% italic(period_begin) + a*","~~italic(r)^2~"="~r2, 
         list(a = format(unname(coef(m)[1]), digits = 2),
              b = format(unname(coef(m)[2]), digits = 2),
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));
}

ggplot(state_market.df, aes(x = as.yearmon(period_begin), y = median_sale_price)) +
  geom_bin_2d(binwidth = c(1/12, 10)) +
  xlab("Time") +
  ylab("Median Sales Price") +
  ggtitle("Median Sales Price vs Time") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_x_yearmon(n = 10) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  annotate("text", x = as.numeric(as.yearmon("2014-06-01")), y = 3000, parse = TRUE, label = lm_eqn(state_market.df))
```

\newpage

Below, we have created 2 boxplots for the types of housing units. One for the median sale price, the other for the median list price. For our outlier tests, we will test if anything is above the 99th quantile. We chose to be less sensitive in our outlier detection, since house prices vary wildly in the market.

#### Both townhouse and multi-family have extreme outliers in the median sales price.

```{r}
ggplot(state_market.df, aes(x = property_type, y = median_sale_price)) +
  geom_boxplot(colour="#003366", fill="#66FFFF", alpha=1/2) +
  xlab("Property Type") +
  ylab("Median Sales Price (in thousands)") +
  ggtitle("Property Type vs Median Sales Price") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r, echo = FALSE}
property_type_outlier_table <- function(df, type, sale_price = TRUE){
  data <- with(df, df[property_type == type, ])
  
  if(sale_price){
    upper_bound <- quantile(data$median_sale_price, 0.99, na.rm = TRUE)
    outlier_indices <- which(data$median_sale_price > upper_bound)
  }
  else{
    upper_bound <- quantile(data$median_list_price, 0.99, na.rm = TRUE)
    outlier_indices <- which(data$median_list_price > upper_bound)
  }
  
  print(table(data$region[outlier_indices]))
}

cat("We can see that for the outliers for multi-family housing the state Hawaii has the 
most outliers. This gives us a good explanation in the reason for the outliers, since 
Hawaii is a small state in the middle of the Pacific serverely limiting the supply of land.")

property_type_outlier_table(state_market.df, type = "Multi-Family")

cat("The same thing is repeated with townhouses, but Connecticut has almost as many as Hawaii
now. There is a possibility that the reasoning behind this is the same as Hawaii with the
limited supply of land, but it also could be that mostly everywhere in Connecticut is close 
to a town or city that sports many amenities.")

property_type_outlier_table(state_market.df, type = "Townhouse")
```

#### Below, we can see that the extreme outliers are not as pronounced with the median list price.

```{r}
ggplot(state_market.df, aes(x = property_type, y = median_list_price)) +
  geom_boxplot(colour="#003366", fill="#66FFFF", alpha=1/2) +
  xlab("Property Type") +
  ylab("Median Sales Price (in thousands)") +
  ggtitle("Property Type vs Median List Price") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r, echo = FALSE}
cat("California leads the US in most outliers for all residential house prices while Columbia
and Hawaii are not far behind. Hawaii has the same reasoning for the listing price as what
was explained above. California's outliers can be explained by the housing markets 
predictions that a huge demand for housing will always be present in the state. Columbia has
a low supply compared to the high demand in the housing market.")

property_type_outlier_table(state_market.df, type = "All Residential", sale_price = FALSE)

cat("New York leads the US in most outliers for condo/co-op house prices. New York will 
obviously lead in most outliers, since most people only live in condo's or cooperative 
housing in the state and the housing markets prediction of ever increasing demands of housing
in places like New York City.")

property_type_outlier_table(state_market.df, type = "Condo/Co-op", sale_price = FALSE)

cat("California and Columbia leads the US in most outliers for single residential house prices.
California is the same as described above. Columbia has a low supply compared to the high demand
in the housing market.")

property_type_outlier_table(state_market.df, type = "Single Family Residential", sale_price = FALSE)
```

\newpage

We created a correlation matrix with our important variables. We expected a strong positive correlation between median sale price and median list price, and an even stronger correlation between median sale price and median price-per-square-foot, but a feature that surprised us was that there was a modest negative correlation between median list price year-on-year and the pending sales year-on-year. We believe we can attribute this to people being less likely to buy a house if the house is more expensive.

```{r}
cor.df <- subset.data.frame(state_market.df, select = c(median_sale_price, median_list_price, median_sale_price_yoy, median_list_price_yoy, median_ppsf, median_list_ppsf, homes_sold_yoy, pending_sales_yoy, new_listings_yoy, inventory_yoy),  drop = FALSE)
cor.table <- cor(cor.df, use="pairwise.complete.obs")

rownames(cor.table) <- c("median sale price", "median list price", "median sale price year-on-year", "median list price year-on-year", "median sale ppsf", "median list ppsf", "homes sold year-old-year", "pending sales year-on-year", "new listings year-on-year", "inventory year-on-year")

colnames(cor.table) <- c("median sale price", "median list price", "median sale price year-on-year", "median list price year-on-year", "median sale ppsf", "median list ppsf", "homes sold year-old-year", "pending sales year-on-year", "new listings year-on-year", "inventory year-on-year")

corrplot(cor.table, type="upper")
```

\newpage

## Change of Plans!

Median sale price was our target variable initially, but we could not find any useful relationships between this variable and the other variables. We found a linear equation to fit the median list price over time, but the r-squared value was too low to for this equation to adequately fit the data. This can attributed to a large variation in sale prices between homes.     Because of the previously unidentified factors in our dataset, we can't always assume that all or most neighborhoods will increase in price at the same rate. We'll now take a closer look at the percentage increase in the number of pending sales per neighborhood. The dataset is too large, so let's pick just one State - California.

Our correlation matrix now gives us much more promising results.

```{r}
cor.df <- subset(state_market.df, state_code == 'CA')

cor.df <- subset.data.frame(cor.df,  select = c(median_sale_price, median_list_price, median_sale_price_yoy, median_list_price_yoy, median_ppsf, median_list_ppsf, homes_sold_yoy, pending_sales_yoy, new_listings_yoy, inventory_yoy,sold_above_list_yoy,avg_sale_to_list_yoy))

cor.table <- cor(cor.df, use="pairwise.complete.obs")

corrplot(cor.table, type="upper")
```

\newpage

## Training our Data

We have decided to go with a linear model with the predictors being inventory_yoy and new_listings_yoy and criterion being pending_sales_yoy. These were select due to there strong correlation they exhibit on each other in the correlation matrix above. From our summary below, we see that both of these variables are statistically significant, since there p-values from the t-test [Pr(>|t|)] is close to zero. We also get an adjusted R^2 of 0.7312 meaning our linear model is of good fit to the data. Interestingly, the inventory_yoy has a negative coefficient, while the new_listings_yoy has a positive coefficient.

```{r}
model1 <- lm(pending_sales_yoy~inventory_yoy+new_listings_yoy, cor.df)
summary(model1)
```

```{r}
model2 <- lm(pending_sales_yoy~inventory_yoy, cor.df)
summary(model2)

model3 <- lm(pending_sales_yoy~new_listings_yoy, cor.df)
summary(model3)
```

```{r}
par(mfrow=c(2,2))
plot(model1)
```

\newpage

## Conclusion

We are going to use linear regression to predict the median sale price of different housing units in 5 years. Since our R^2 is low in the binplot equation, we know that only 12.5% of the  independent variable (median sales price) can be explained through time. We can see that this is a small, but reliable relationship. For us to predict future house prices accurately, we will need to account for other factors that are outside the scope of this dataset; however, we may be able to predict a lower or upper bound or a relative area that future house prices might fall into. Either way, we can confidently confirm that the future housing market will be costlier that the current one.

\newpage

## Sample Data

```{r}
sample.df <- subset.data.frame(state_market.df, select = c(period_begin, state_code, property_type, median_sale_price, median_list_price, median_sale_price_yoy, median_list_price_yoy, median_ppsf, median_list_ppsf, homes_sold_yoy, pending_sales_yoy, new_listings_yoy, inventory_yoy),  drop = FALSE)

pander(head(sample.df))
```

